---
title: "Predicting exercise type using classification approach"
author: "Przemyslaw Zientala"
date: "28 February 2016"
output: html_document
---

#Introduction
The aim of this project is to predict (*classify*) the manner in which subjects of [a study](http://groupware.les.inf.puc-rio.br/har) exercised. To accomplish this, a model selection procedure was utilised and the best model was selected according to its accuracy.

#Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal was to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website [here](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).

#Analysis

##Data cleaning
Since both datasets contained many redundant columns containing little to no information (either "NA" vlaues or empty cells), these were removed to make the analysis and model fitting easier.

```{r, cache=TRUE}
#Load the data
training <- read.csv("pml-training.csv")
testing <- read.csv("pml-testing.csv")

#Select complete columns
na.prop <- function(x){sum(is.na(x))/length(x)*100}
cols_prop <- sapply(training, na.prop)
col_names <- names(cols_prop[(cols_prop < 10)])

#Check number of "NA" values
sum(is.na(training[,col_names]))

training <- training[,col_names]
```

As it can be seen, there are no "NA" values left, which is a good sign.

Now, let's remove all empty values:
```{r, cache=TRUE}
#Select columns with no empty values (not the same as NA!)
prop.empty <- function(x){sum(x == "")/length(x)*100}
cols_prop <- sapply(training, prop.empty)
col_names <- names(cols_prop[cols_prop < 10])

#Check the number of empty cells
sum((sapply(training[,col_names], prop.empty) > 0))
training <- training[,col_names]
```

Again, there are no empty values left, so let's proceed to clean testing dataset using the same approach:

```{r, cache=TRUE}
#Select complete columns
na.prop <- function(x){sum(is.na(x))/length(x)*100}
cols_prop <- sapply(testing, na.prop)
col_names <- names(cols_prop[(cols_prop < 10)])

#Check number of "NA" values
sum(is.na(testing[,col_names]))

testing <- testing[,col_names]

#Select columns with no empty values (not the same as NA!)
prop.empty <- function(x){sum(x == "")/length(x)*100}
cols_prop <- sapply(testing, prop.empty)
col_names <- names(cols_prop[cols_prop < 10])

#Check the number of empty cells
sum((sapply(testing[,col_names], prop.empty) > 0))
testing <- testing[,col_names]

#Remove "X" col
training <- training[,-c(1:5)]
testing <- testing[,-c(1:5)]
```

##Splitting the data
Now, all the data is ready to be split into training and validation sets. To ensure reproducibility, the seed was set to 1:
```{r}
#Split data into train and test sets
set.seed(1)
train_rows <- sample(1:nrow(training), size = 2/3*nrow(training))
train_data <- training[train_rows,]
test_data <- training[which(1:nrow(training) %in% train_rows == FALSE),]
```

Now, let's proceed to model and variable selection.

#Exploratory data analysis
It would be 